\section{Changes}

\subsection{Version \version}

\subsection{Version 1.5-4}

\begin{itemize}
\item Fix typos.
\item \changeinsection{expressions}{fix syntax of guards in iterators}
\item \changeinsection{semantics}{fix definition of undefined terms and
  predicates}
\item \changeinsection{typing}{no user-defined types}
\item \changeinsection{builtinconstructs}{no more implementation issue for
  \lstinline|\\old|}
\item \changeinsection{at}{more restrictive scoping rule for label references 
in \lstinline|\\at|}
\end{itemize}

\subsection{Version 1.5-3}

\begin{itemize}
\item Fix various typos.
\item Warn about features known to be difficult to implement.
\item \changeinsection{expressions}{fix semantics of ternary operator}
\item \changeinsection{expressions}{fix semantics of cast operator}
\item \changeinsection{expressions}{improve syntax of iterator quantifications}
\item \changeinsection{semantics}{improve and fix example \ref{ex:semantics}}
\item \changeinsection{loop_annot}{improve explanations about loop invariants}
\item \changeinsection{logicalstates}{add hybrid functions and predicates}
\end{itemize}

\subsection{Version 1.5-2}

\begin{itemize}
\item \changeinsection{expressions}{remove laziness of operator
  \lstinline|<==>|}
\item \changeinsection{expressions}{restrict guarded quantifications to integer}
\item \changeinsection{expressions}{add iterator quantifications} 
\item \changeinsection{expressions}{extend unguarded quantifications to char}
\item \changeinsection{locations}{extend syntax of set comprehensions}
\item \changeinsection{loop_annot}{simplify explanations for loop invariants and
  add example.}
\end{itemize}

\subsection{Version 1.5-1}

\begin{itemize}
\item Fix many typos.
\item Highlight constructs with semantic changes in grammars.
\item Explain why unsupported features have been removed.
\item Indicate that experimental \acsl features are unsupported.
\item Add operations over memory like \lstinline|\valid|.
\item \changeinsection{expressions}{lazy operators \lstinline|&&|,
  \lstinline+||+, \lstinline|\^\^|, \lstinline|==>| and \lstinline|<==>|}
\item \changeinsection{expressions}{allow unguarded quantification over boolean}
\item \changeinsection{expressions}{revise syntax of \lstinline|\\exists|}
\item \changeinsection{semantics}{better semantics for undefinedness}
\item \changeinsection{locations}{revise syntax of set comprehensions}
\item \changeinsection{loop_annot}{add loop invariants, but they lose their
  inductive \acsl nature}
\item \changeinsection{generalmeasures}{add general measures for termination}
\item \changeinsection{specmodules}{add specification modules}
\end{itemize}

\subsection{Version 1.5-0}

\begin{itemize}
\item Initial version.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\ifthenelse%
    {\boolean{PrintImplementationRq}}%
    {
\section{Changes in \eacsl implementation}

\subsection{Version \eacslversion}

\begin{itemize}
\item Implement bitwise complementation.
\end{itemize}

\subsection{Version 0.1}

\begin{itemize}
\item Initial version.
\end{itemize}

    }%
    {}
