\subsection{Version \version}
\begin{itemize}
\item Remove experimental and barely used Logic attribute annotations.
Use standard C attributes instead.
\end{itemize}
\subsection{Version 1.21}
\begin{itemize}
\item More precise definition of pre- and post-state of a function contract
  (section~\ref{sec:pre-post-state})
\item Extend grammar for module definitions and allow importing external symbols
  (section~\ref{sec:specmodules})
\item Set the default value for missing \lstinline|terminates| to
  \lstinline|\false| if the attribute \lstinline|_NoReturn| is present
  (section~\ref{sec:non-term-funct}).
\end{itemize}
\subsection{Version 1.20}
\begin{itemize}
\item Fix the default value for missing \lstinline|exits| (to
\lstinline|\false|) (section~\ref{sec:multiplecontracts}).
\end{itemize}
\subsection{Version 1.19}
\begin{itemize}
\item Introduces built-in predicates \lstinline|\object_pointer| and
\lstinline|\pointer_comparable| (section~\ref{subsec:memory}).
\end{itemize}
\subsection{Version 1.18}
\begin{itemize}
\item Clarifies meaning of \lstinline|terminates| and
  \lstinline|decreases|/\lstinline|variant| clauses (section~\ref{sec:non-term-funct})
\end{itemize}
\subsection{Version 1.17}
\begin{itemize}
\item explicit role of \lstinline|check| and \lstinline|admit| \lstinline|requires|
clause with respect to \lstinline|complete| and \lstinline|disjoint| clauses
(section~\ref{sec:compl-behav})
\end{itemize}
\subsection{Version 1.16}
\begin{itemize}
\item Slightly improve section on logic attribute annotations
(removed in Version \version), that was a bit rushed into
the last version
\item Introduce \lstinline|check| and \lstinline|admit| clause kinds
(sections~\ref{sec:check-admit-clauses}, \ref{sec:assertions},
\ref{sec:loop_annot},~\ref{sec:generalized-invariants}, and~\ref{sec:lemmas}).
\end{itemize}
\subsection{Version 1.15}
\begin{itemize}
\item Add section~\ref{sec:ppimpl} for precising status of pre-processing
  directives and macros against specifications
\item Introduction of the \lstinline|\ghost| qualifier (section~\ref{sec:ghost})
\end{itemize}
\subsection{Version 1.14} % Potassium and Calcium
\begin{itemize}
\item Introduce \lstinline|check| annotation (section~\ref{sec:assertions})
\end{itemize}

\subsection{Version 1.13} % Chlorine and Argon
\begin{itemize}
\item New infix predicate \lstinline|\in| for set membership
  (section~\ref{sec:locations})
\item Fixes some typing error for constructs rejecting
  \lstinline|void*| pointers (section~\ref{subsec:allocation-contract})
\item Notations added for real numbers $\pi$ and $e$
  (section~\ref{sec:floating-point})
\end{itemize}

\subsection{Version 1.12} % Sulfur
\begin{itemize}
\item Fixes syntax rule for statement contracts in allowing completeness clauses
  (figure~\ref{fig:gram:stcontracts})
\end{itemize}

\subsection{Version 1.11}
\begin{itemize}
\item Functions related to infinites and the sign of floating-point value
  (section~\ref{sec:floating-point})
\item New section for predicates related to well-typedness
  (section~\ref{sec:welltyped})
\item Syntax for defining a set by giving explicitely its elements
  (section~\ref{sec:locations})
\item Adding lists as first-class values (section~\ref{sec:lists})
\item Change the associativity of bitwise operator \lstinline|-->| to right,
  in accordance with the one of \lstinline|==>| operator
\item Glyph used for \lstinline|^^| operator (xor) fixed
\end{itemize}

\subsection{Version 1.10}
\begin{itemize}
\item Change keyword for importing libraries (section~\ref{sec:specmodules})
\item Fix numerous typos reported by David Cok
\item Disallow meaningless \lstinline|assigns \nothing \from x|
  (section~\ref{sec:func-dep})
\end{itemize}

\subsection{Version 1.9}
\begin{itemize}
\item Fix typo in definition of \lstinline|\fresh| predicate
  (section~\ref{sec:allocation-clause})
\item Fix grammar inconsistencies
  \begin{itemize}
  \item use proper C rules names
  \item fix mismatch in non-terminal names
  \end{itemize}
\item Rename "Unspecified values" to "Dangling pointers" and precise it
  (section~\ref{sec:dangling_pointers})
\end{itemize}

\subsection{Version 1.8}
\begin{itemize}
\item Mention binary literal constant typing
\end{itemize}

\subsection{Version 1.7}
\begin{itemize}
\item Added missing shift operators in figure~\ref{fig:gram:term}
\item Modified syntax for naming terms and predicates (figures~\ref{fig:gram:pred} and \ref{fig:gram:term})
\item Added syntax rule for literal constants (figure~\ref{fig:gram:term})
\end{itemize}

\subsection{Version 1.6}
\begin{itemize}
\item Modified syntax for model fields (section~\ref{sec:model})
\item Added missing logical xor operator (figure~\ref{fig:gram:term}).
\item Addition of logical labels related to loops
  (section~\ref{sec:default-logic-labels}).
\item Addition of labels to built-ins related to memory blocks
  (section~\ref{subsec:memory})
\item Introduction of \lstinline|\valid_read| built-in and clarification of
  the notion of validity (section~\ref{subsec:memory}).
\item Introduction of built-in \lstinline|\allocable|,
  \lstinline|\allocation|, \lstinline|\freeable| and
  \lstinline|\fresh| (section~\ref{sec:allocation-clause}).
\item Introduction of \allocates and \frees clauses
  (section~\ref{sec:allocation-clause}).
\item Clarify the semantics of \assigns clauses into statement contract.
\item Improvements to the \volatile clause
  (section~\ref{sec:volatile-variables}).
\item Clarify the evaluation of arrays inside an \lstinline|at|
  (section~\ref{sec:at}).
\end{itemize}

\subsection{Version 1.5}

\begin{itemize}
\item Clarify the status of \Loop \invariant in presence of
  \Break or side-effects in the loop test.
\item Introduction of \lstinline|\with| keyword for functional updates.
\item Added bnf entry for completeness of function behaviors.
\item Order of clauses in statement contracts is now fixed.
\item \requires clauses are allowed before behaviors of statement contracts.
\item Added explicit singleton construct for sets.
\item Introduction of logical arrays.
\item Operations over pointers and arrays have been precised.
\item Predicate \lstinline|\initialized| (section~\ref{sec:initialized})
  now takes a set of pointers as argument.
\end{itemize}

\subsection{Version 1.4}
\begin{itemize}
\item Added UTF-8 counterparts for built-in types
  (\lstinline|integer|, \lstinline|real|, \lstinline|boolean|).
\item Fixed typos in the examples corresponding to features
  implemented in Frama-C.
\item Order of clauses in function contracts is now fixed.
\item Introduction of abrupt termination clauses.
\item Introduction of \lstinline|axiomatic| to gather predicates, logic
  functions, and their defining axioms.
\item Added specification templates appendix for common specification issues.
\item Use of sets as first-class term has been precised.
\item Fixed semantics of predicate \lstinline|\separated|.
\end{itemize}

\subsection{Version 1.3}
\begin{itemize}
\item Functional update of structures.
\item Terminates clause in function behaviors.
\item Typos reported by David Mentr\'e.
\end{itemize}

\subsection{Version 1.2}
This is the first public release of this document.
