\section{Comparison with JML}
\label{sec:comp-jml}

Although we took our inspiration in the Java Modeling Language (aka
JML~\cite{jmlhomepage}), ACSL is notably different from JML in
two crucial aspects:

\begin{itemize}
\item ACSL is a BISL for C, a low-level structured language, while JML
  is a BISL for Java, an object-oriented inheritance-based high-level
  language. Not only the language features are not the same but the
  programming styles and idioms are very different, which entails also
  different ways of specifying behaviors. In particular, C has no
  inheritance nor exceptions, and no language support for the simplest
  properties on memory (e.g., the size of an allocated memory block).
\item JML relies on runtime assertion checking (RAC) when typing,
  static analysis and automatic deductive verification fail. The
  example of CCured~\cite{necula02ccured,condit03ccured}, that adds
  strong typing to C by relying on RAC too, shows that it is not possible
  to do it in a modular way. Indeed, it is necessary to modify the
  layout of C data structures for RAC, which is not modular. The
  follow-up project Deputy~\cite{condit07deputy} thus reduces the
  checking power of annotations in order to preserve modularity.  On
  the contrary, we choose not to restrain the power of annotations
  (e.g., all first order logic formulas are allowed). To that end, we
  rely on manual deductive verification using an interactive theorem
  prover (e.g., Coq) when every other technique failed.
\end{itemize}

\noindent
In the remainder of this chapter, we describe these differences in
further details.

\subsection{Low-level language vs. inheritance-based one}

\subsubsection*{No inherited specifications}

JML has a core notion of inheritance of specifications, that duplicates
in specifications the inheritance feature of Java.  Inheritance
combined with visibility and modularity account for a number of
complex features in JML (e.g., \verb|spec_public| modifier, data
groups, represents clauses, etc), that are necessary to express the
desired inheritance-related specifications while respecting visibility
and modularity. Since C has no inheritance, these intricacies are
avoided in ACSL.

\subsubsection*{Error handling without exceptions}
\label{sec:errorhandling}

The usual way of signaling errors in Java is through
exceptions. Therefore, JML specifications are tailored to express
exceptional postconditions, depending on the exception raised. Since C
has no exceptions, ACSL does not use exceptional
specifications. Instead, C programmers are used to signal errors by
returning special values, like mandated in various ways in the C standard.

\begin{example}
In \S 7.12.1 of the standard, it is said that functions in <math.h>
signal errors as follows:
``On a domain error, [...] the integer expression errno
acquires the value EDOM.''
\end{example}

\begin{example}
In \S 7.19.5.1 of the standard, it is said that function fclose signals
errors as follows:
``The fclose function returns [...] EOF if any errors were detected.''
\end{example}

\begin{example}
In \S 7.19.6.1 of the standard, it is said that function fprintf
signals errors as follows:
``The fprintf function returns [...] a negative value if an output or
encoding error occured.''
\end{example}

\begin{example}
In \S 7.20.3 of the standard, it is said that memory management functions
signal errors as follows:
``If the space cannot be allocated, a null pointer is returned.''
\end{example}

As shown by these few examples, there is no unique way to signal
errors in the C standard library, not mentioning user-defined
functions. But since errors are signaled by returning special values, it
is sufficient to write an appropriate postcondition:

\begin{flushleft}\ttfamily
/*@ ensures $\mathit{\result == error\_value ~||~ normal\_postcondition}$; */
\end{flushleft}

% \noindent
% A tool could easily set error conditions aside, by providing an
% appropriate extension of behaviors, e.g., using a new keyword
% \verb|failswith|:

% \begin{flushleft}\ttfamily
% /*@ failswith $\mathit{\result == error\_value}$; \\
% ~~@ ensures $\mathit{normal\_postcondition}$; \\
% ~~@*/
% \end{flushleft}


%\input{fwrite-malloc.pp}


\subsubsection*{C contracts are not Java ones}

In Java, the precondition of the following function that nullifies an
array of characters is always true. Even if there was a precondition
on the length of array {\ttfamily a}, it could easily be expressed using
the Java expression {\ttfamily a.length} that gives the dynamic length
of array {\ttfamily a}.

\begin{flushleft}\ttfamily
public static void Java\_nullify(char[] a) \{ \\
~~if (a == null) return; \\
~~for (int i = 0; i < a.length; ++i) \{ \\
~~~~a[i] = 0; \\
~~\} \\
\}
\end{flushleft}

On the contrary, the precondition of the same function in C, whose
definition follows, is more involved. First, remark that the
C programmer has to add an extra argument for the size of the array,
or rather a lower bound on this array size.

\begin{flushleft}\ttfamily
void C\_nullify(char* a, unsigned int n) \{ \\
~~int i; \\
~~if (n == 0) return; \\
~~for (i = 0; i < n; ++i) \{ \\
~~~~a[i] = 0; \\
~~\} \\
\}
\end{flushleft}

\noindent
A correct precondition for this function is the following:

\begin{flushleft}\ttfamily
/*@ requires $\mathit{\valid(a + 0..(n-1))}$; */
\end{flushleft}

where predicate $\valid$ is the one defined in Section~\ref{subsec:memory}.
(note that $\valid(a + 0..(-1))$ is the same as
$\valid(\verb+\empty+)$ and thus is true regardless of the validity of
$a$ itself).
When {\ttfamily n} is null, {\ttfamily a} does
not need to be valid at all, and when {\ttfamily n} is strictly
positive, {\ttfamily a} must point to an array of size at least
{\ttfamily n}.  To make it more obvious, the C programmer adopted a
defensive programming style, which returns immediately when {\ttfamily n} is
null. We can duplicate this in the specification:

\begin{flushleft}\ttfamily
/*@ requires $\mathit{n == 0 ~||~ \valid(a + 0..(n-1))}$; */
\end{flushleft}

Usually, many memory requirements are only necessary for some paths
through the function, which correspond to some particular
behaviors, selected according to some tests performed along the
corresponding paths. Since C has no memory
primitives, these tests involve other variables that the C programmer
added to track additional information, like {\ttfamily n} in our example.

To make it easier, it is possible in ACSL to distinguish between the
\verb|assume| part of a behavior, that specifies the tests that need
to succeed for this behavior to apply, and the \verb|requires| part
that specifies the additional preconditions that must be true when a
behavior applies. The specification for our example can then be
translated into:

\begin{flushleft}\ttfamily
/*@ behavior $\mathit{n\_is\_null}$: \\
~~@ ~~assumes $\mathit{n == 0}$; \\
~~@ behavior $\mathit{n\_is\_not\_null}$: \\
~~@ ~~assumes $\mathit{n ~> 0}$; \\
~~@ ~~requires $\mathit{\valid(a + 0..(n-1))}$; \\
~~@*/
\end{flushleft}

This is equivalent to the previous requirement, except here behaviors
can be completed with postconditions that belong to one behavior only.
Contrary to JML, the set of behaviors for a function do not
necessarily cover all cases of use for this function, as mentioned in
Section~\ref{subsec:behaviors}. This allows for partial
specifications, whereas JML behaviors cannot offer such
flexibility. Here, Our two behaviors are clearly mutually exclusive,
and, since \texttt{n} is an \texttt{unsigned int}, our
they cover all the possible cases. We could have specified that as well, by
adding the following lines in the contract (see
Section~\ref{sec:compl-behav}).
\begin{flushleft}\ttfamily
/*@~~...\\
~~@~~disjoint behaviors;\\
~~@~~complete behaviors;\\
~~@*/
\end{flushleft}

\subsubsection*{ACSL contracts vs. JML ones}

To fully understand the difference between specifications in ACSL and
JML, we detail in below the requirement on the pre-state and
the guarantee on the post-state given by behaviors in JML and ACSL.

A JML contract is either \emph{lightweight} or \emph{heavyweight}.
For the purpose of our comparison, it is sufficient to know that a
lightweight contract has \verb|requires| and \verb|ensures| clauses
all at the same level, while an heavyweight contract has multiple
behaviors, each consisting of \verb|requires| and \verb|ensures|
clauses. Although it is not possible in JML to mix both styles, we can
define here what it would mean to have both, by conjoining the
conditions on the pre- and the post-state.
Here is an hypothetical JML contract mixing lightweight and
heavyweight styles:
\begin{flushleft}\ttfamily
/*@ requires $P_1$; \\
~~@ requires $P_2$; \\
~~@ ensures  $Q_1$; \\
~~@ ensures  $Q_2$; \\
~~@ behavior $x_1$: \\
~~@ ~~requires $A_1$; \\
~~@ ~~requires $R_1$; \\
~~@ ~~ensures $E_1$; \\
~~@ behavior $x_2$: \\
~~@ ~~requires $A_2$; \\
~~@ ~~requires $R_2$; \\
~~@ ~~ensures $E_2$; \\
~~@*/
\end{flushleft}
It assumes from the pre-state the condition:
\begin{center}
{\ttfamily $P_1 ~\&\&~ P_2 ~\&\&~ ((A_1 ~\&\&~ R_1) ~||~ (A_2 ~\&\&~ R_2))$}
\end{center}
and guarantees that the following condition holds in post-state:
\begin{center}
{\ttfamily $Q_1 ~\&\&~ Q_2 ~\&\&~
  (\old(A_1 ~\&\&~ R_1) ==> E_1) ~\&\&~ (\old(A_2 ~\&\&~ R_2) ==> E_2)$}
\end{center}


Here is now an ACSL specification:

\begin{flushleft}\ttfamily
/*@ requires $P_1$; \\
~~@ requires $P_2$; \\
~~@ ensures  $Q_1$; \\
~~@ ensures  $Q_2$; \\
~~@ behavior $x_1$: \\
~~@ ~~assumes $A_1$; \\
~~@ ~~requires $R_1$; \\
~~@ ~~ensures $E_1$; \\
~~@ behavior $x_2$: \\
~~@ ~~assumes $A_2$; \\
~~@ ~~requires $R_2$; \\
~~@ ~~ensures $E_2$; \\
~~@*/
\end{flushleft}

\noindent
Syntactically, the only difference with the JML specification is the
addition of the \verb|assumes| clauses.
Its translation to assume-guarantee is however quite different.
It assumes from the pre-state the condition:

\begin{center}
{\ttfamily $P_1 ~\&\&~ P_2 ~\&\&~
  (A_1 ==> R_1) ~\&\&~ (A_2 ==> R_2)$}
\end{center}

\noindent
and guarantees that the following condition holds in the post-state:

\begin{center}
{\ttfamily $Q_1 ~\&\&~ Q_2 ~\&\&~
  (\old(A_1) ==> E_1) ~\&\&~ (\old(A_2) ==> E_2)$}
\end{center}

Thus, ACSL allows to distinguish between the clauses that control
which behavior is active (the \texttt{assumes} clauses) and the
clauses that are preconditions for a particular behavior (the internal
\texttt{requires} clauses). In addition, as mentioned above, there is
by default no requirement in ACSL for the specification to be complete (The
last part of the JML condition on the pre-state). If desired, this has
to be precised explicitely with a \verb+complete behaviors+ clause as
seen in Section~\ref{sec:compl-behav}.

\subsubsection*{ACSL \texttt{assigns} clause}

Beware that in \texttt{assigns} clauses, the given locations refer to
the post-state, whereas in JML, they refer to the pre-state. The JML
semantics can be obtained using \verb|\old| construct.

\subsection{Deductive verification vs. RAC}

\subsubsection*{Sugar-free behaviors}

As explained in details in~\cite{raghavan00desugaring}, JML
heavyweight behaviors can be viewed as syntactic sugar (however
complex it is) that can be translated automatically into more basic
contracts consisting mostly of pre- and postconditions and frame
conditions.  This allows complex nesting of behaviors from the user
point of view, while tools only have to deal with basic contracts. In
particular, the major tools on JML use this desugaring process, like
the Common JML tools to do assertion checking, unit testing,
etc. (see~\cite{leavens00jml}) and the tool ESC/Java2 for
automatic deductive verification of JML specifications
(see~\cite{kiriny07implnotes}).

One issue with such a desugaring approach is the complexity of the
transformations involved, as \emph{e.g.} for desugaring assignable clauses
between multiple \textit{spec-cases} in
JML~\cite{raghavan00desugaring}.  Another issue is precisely that
tools only see one global contract, instead of multiple independent
behaviors, that could be analyzed separately in more detail.
Instead, we favor the view that a function implements multiple
behaviors, that can be analyzed separately if a tool feels like
it. Therefore, we do not intend to provide a desugaring process.

\subsubsection*{Axiomatized functions in specifications}

JML only allows pure Java methods to be called in
specifications~\cite{leavens00preliminary}. This is certainly
understandable when relying on RAC: methods called should be defined
so that the runtime can call them, and they should not have
side-effects in order not to pollute the program they are supposed to
annotate.

In our setting, it is desirable to allow calls to logical functions in
specifications. These functions may be defined, like program ones, but
they may also be only declared (with a suitable declaration of \reads
clause) and defined through an axiomatization.
This makes for richer specifications that may be useful either in
automatic or in manual deductive verification.

\subsection{Syntactic differences}

The following table summarizes the difference between JML keywords and
ACSL ones, when the intent is the same, although minor differences
might exist.

\begin{center}
\begin{tabular}{|l|l|}
\hline
  JML                  & ACSL \\ \hline
  modifiable,assignable           & assigns \\
  measured\_by         & decreases \\
  loop\_invariant      & loop invariant \\
  decreases            & loop variant \\
  $(\Forall~\tau~x ; P ; Q)$ &   $(\Forall~\tau~x ; P \verb|==>| Q)$ \\
  $(\Exists~\tau~x ; P ; Q)$ &   $(\Exists~\tau~x ; P \verb|&&| Q)$ \\
  $(\verb|\max|~\tau~x ; a \verb|<=| x \verb|<=| b ; f)$ &
  $\verb|\max|(a,b,\verb|\lambda|~\tau~x ; f)$ \\
  \hline
\end{tabular}
\end{center}


\oldremark{PC}{

  ne pas diverger de JML quand ce n'est pas justifie'


  -> rediscuter du decreases versus loop variant

}


%%% Local Variables:
%%% mode: latex
%%% TeX-PDF-mode: t
%%% TeX-master: "main"
%%% End:
